---
title: "(Post-)Preprocessing of proteomics and RNA"
author: "Fiona Dick"
date: "`r Sys.Date()`"
output: html_document
	toc: true
runtime: shiny
---

```{r}
library(boot)
library(mixOmics)
library(ggbiplot)
library(tibble)
library(ermineR)
library(ggpubr)
library("DESeq2")
require("ggplot2")
require("kableExtra")
#install.packages("devtools")
#devtools::install_github("rlbarter/superheat")
#library(superheat)
require("ggfortify")
require("readxl")
#library(PathCluster)
require("magrittr")
require("gridExtra")
#library("shinythemes")
library("dbscan")
require("proteus")
require("dplyr")
require("tidyr")
require("preprocessCore")         
require("readr")
require("knitr")
require("kableExtra")
require("ggpubr")
require("stringr")
require("pheatmap")
require("RColorBrewer")
require("ggrepel")
require("patchwork")
require("ggplotify")
library("magrittr")
library("DRIMSeq")
library("DEXSeq")
library("DTU")
library(ComplexHeatmap)
library(viridis)
library(reshape2)
library(markerGeneProfile)
library("homologene")
library("markerGeneProfile")
data("mouseMarkerGenesCombined")
```

##  Utils

```{r}
source("./functions.R")
```

```{r}
cols <- c("#ef476f", "#073b4c", "#118ab2", "#06d6a0", "#84a59d")
names(cols) <- c("PD", "HA", "Other", "YG", "Reference") 
```


## Paths

```{r}
# RNA
gtfPath  <- "./referenceData/gencode.v32.annotation.gtf"
#remember to uncompress before using
salmonPath  <- "./salmonOut"
```


## Read metadata

```{r}
info_rna <- readRDS("./rdsData/info_rna.rds") 
info_rna_pd <- readRDS("./rdsData/info_rna_pd.rds")
#proteomics experiment 
info <- readRDS("./rdsData/info.rds")
```

## Load and normalize proteomics data 

From the ProteinGroups (maxQuant output, not provided) file create three count matrices:  
* raw from maxQuant output (loaded from rds (rawData.rds)
* remove zeros
* with quantile normalization 
* with quantile normalization followed by batch correction
* normalized with the TMT normalization function of Proteus
* normalized with TMT norm. func. of Proteus followed by batch correction


```{r}
rawData <- readRDS("./rdsData/rawData.rds")
```

```{r}
# determine proteins that are making up 5% of the sum of all protein intensities in more than n samples
topProts <- investigateTopGenes(rawData, name_cols = c("Gene.names", "Protein.IDs"))[[1]][,"Protein.IDs"]
# remove extremly abundant proteins and zeros
rawData_filt  <- rawData %>%
	dplyr::mutate_at(info$reporter.intensity.id, as.numeric) %>%
	dplyr::filter(!(Protein.IDs %in% topProts)) %>%
	dplyr::filter_if(is.numeric, all_vars(. != 0))

# Alternative to removing all zeros:
# Removing proteins that have intensity == 0 in **all** samples within a batch,
# in at least **n** batches
# rawData_filt <- remove_zeros(rawData_filt, info, n = 2)

# Normalize options
m_quant  <- rawData_filt %>% loading_norm(., info, method = "quantile", within_batch = T, use_median = T, name_cols = c("Gene.names", "Protein.IDs")) 
rownames(m_quant) <- m_quant$Protein.IDs

m_lib <-  rawData_filt %>% loading_norm(., info, method = "lib_size", within_batch = T, use_median = T, name_cols = c("Gene.names", "Protein.IDs"))  
rownames(m_lib) <- m_lib$Protein.IDs

m_logLib <-  rawData_filt %>% loading_norm(., info, method = "log_lib_size", within_batch = T, use_median = T, name_cols = c("Gene.names", "Protein.IDs")) 
rownames(m_lib) <- m_lib$Proten.IDs

# Apply batch correction 
m_lib_batch <- batch_corr(m_lib, info)
rownames(m_lib) <- m_lib$Protein.IDs

# Apply batch correction 
m_batch <- batch_corr(rawData_filt, info)
rownames(m_batch) <- m_batch$Protein.IDs

# Apply batch correction followed by quantile correction
m_batch <- batch_corr(rawData_filt, info)
m_batch_norm <- loading_norm(m_batch, info, method = "log_lib_size")
rownames(m_batch_norm) <- m_batch_norm$Protein.IDs

# Apply batch correction followed by scaling
m_batch <- batch_corr(rawData_filt, info)
# dividing by root mean square sqrt(sum(x^2)/(n-1)), keep center == F 
m_batch_scld <- as.data.frame(scale(m_batch[,-c(1,2)], center = F))
m_batch_scld$Protein.IDs  <- m_batch$Protein.IDs
m_batch_scld$Gene.names  <- m_batch$Gene.names
rownames(m_batch_scld) <- m_batch_scld$Protein.IDs
```

### QC Suppl. Figure 1


```{r}
rownames(rawData_filt)  <- rawData_filt$Protein.IDs
rawData_filt %>% dplyr::select(-Gene.names, -Protein.IDs) -> df
rownames(info) <- info$reporter.intensity.id
ggbiplot::ggbiplot(prcomp(t(df), scale = T),
		   groups = info[colnames(df),"batch"],
		   varname.size = 0, size = 1,
		   var.axes = F, ellipse = T) +
	  cowplot::theme_cowplot()  +
	  labs(tag = "A", col = "Groups") +
	  theme(legend.position = "None") -> p1
rownames(m_quant)  <- m_quant$Protein.IDs
m_quant %>% dplyr::select(-Gene.names, -Protein.IDs) -> df
rownames(info) <- info$reporter.intensity.id
ggbiplot::ggbiplot(prcomp(t(df), scale = T),
		   groups = info[colnames(df),"batch"],
		   varname.size = 0, size = 1,
		   var.axes = F, ellipse = T) +
	  cowplot::theme_cowplot()  +
	  labs(col = "Groups", tag = "B") -> p1.1
rownames(m_batch_scld)  <- m_batch_scld$Protein.IDs
m_batch_scld %>% dplyr::select(-Gene.names, -Protein.IDs) -> df
rownames(info) <- info$reporter.intensity.id
ggbiplot::ggbiplot(prcomp(t(df), scale = T),
		   groups = info[colnames(df),"batch"],
		   varname.size = 0, size = 1,
		   var.axes = F, ellipse = F) +
	  cowplot::theme_cowplot()  +
	  labs(col = "Groups", tag = "C") +
	  theme(legend.position = "bottom") -> p2
ggbiplot::ggbiplot(prcomp(t(df), scale = T),
		   groups = info[colnames(df),"condition1"],
		   varname.size = 0, size = 1,
		   var.axes = F, ellipse = F) +
	  cowplot::theme_cowplot()  +
	  labs(col = "Groups", tag = "D") +
	  scale_color_manual(values = cols) +
	  theme(legend.position = "bottom") -> p3
  
pdf("./result_figs/Supp/proteomics_qc/SFig1.pdf", width = 10, height = 10)
(p1 + p1.1 ) / (p2 + p3)
dev.off()
```


## RNA for HA and YG samples

```{r gene-levelCPM}
# gene-level counts
# create gene-tx annotation from ucsc hg19 gtf
annot_general <- get_annot("86", gtf_path = gtfPath, strip_version = F)
tx2gene <- annot_general[,c(2,1)]
colnames(tx2gene) <- c("tx","gene")
# salmon raw data dirs
files_per_cohort <- file.path(salmonPath, info_rna$Norway$sample_id, "quant.sf")
names(files_per_cohort)  <- info_rna$Norway$sample_id
# import summerized to gene-level (txOut=F)
cpmAll <- import_and_filter(files_per_cohort, info_rna$Norway$sample_id, tx2gene, gtfPath)
gene_symbols <- cpmAll %>%
	dplyr::select(gene_id, gene_name)
```



## RNA for PD samples

Raw RNA from salmon Dirs and sample ids from RNASeq experiment info file

cpmAll_pd

```{r}
# Included  the study
annot_general <- get_annot("86", gtf_path = gtfPath, strip_version = F)
tx2gene <- annot_general[,c(2,1)]
colnames(tx2gene) <- c("tx","gene")
# salmon raw data dirs
files_per_cohort <- file.path(salmonPath, info_rna_pd$sample_id, "quant.sf")
names(files_per_cohort)  <-  info_rna_pd$sample_id
cpmAll_pd <- import_and_filter(files_per_cohort, info_rna_pd$sample_id, tx2gene, gtfPath)
```


## Saving preprocessed data for paper code 

```{r}
saveRDS(cpmAll, "../../rna_protein_coupling_PD/rdsData/cpmAll.rds")
saveRDS(cpmAll_pd, "../../rna_protein_coupling_PD/rdsData/cpmAll.rds")
saveRDS(rawData_filt, "../../rna_protein_coupling_PD/rdsData/rawData_filt.rds")
saveRDS(m_quant, "../../rna_protein_coupling_PD/rdsData/m_quant.rds")
saveRDS(m_lib, "../../rna_protein_coupling_PD/rdsData/m_lib.rds")
saveRDS(m_logLib, "../../rna_protein_coupling_PD/rdsData/logLib.rds")
saveRDS(m_batch, "../../rna_protein_coupling_PD/rdsData/m_batch.rds")
saveRDS(m_lib_batch, "../../rna_protein_coupling_PD/rdsData/m_lib_batch.rds")
saveRDS(m_batch_scld, "../../rna_protein_coupling_PD/rdsData/m_batch_scld.rds")
```



